---
title: "index.qmd"
author: "Kitiibwa Melita Nathania"
date: "12/11/2025"
format: 
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
editor: visual
editor_options:
  chunk_output_type: console
---

# Introduction

This data was obtained from <https://www.waterpointdata.org/2021/11/09/global-water-challenge-announces-the-development-of-whdx-an-open-data-exchange-to-improve-wash-in-healthcare-facilities/>

under the aquaya waterTracs program.

private information such as ids were removed. in the dictionary file you will fing important cariables that may need some interpretation.

**PROJECT DESCRIPTION**

To identify the organisations that constructed the different water points

To assess their safety as drinking water points

To identify who manages the water points.

## Methods

Involved renaming a few columns as well as mutating some to give specific values. The Na values could not be removed as they would make the results meaningless.

```{r}
library(tidyverse)
library(here)

```

```{r}
raw_data <- read_csv(here::here("data/raw/CAPSTONE.csv"))
```

```{r}
raw_data_new <- raw_data |> 
  rename(year = install_year,
        fecal_value = fecal_coliform_value,
         fecal_presence = fecal_coliform_presence,
         country_id = clean_country_id,
         country_name = clean_country_name,
         adm1 = clean_adm1,
         tech = water_tech_clean,
         management = management_clean,
         status = status_clean
  )
```

raw_new_data is the processed folder being used.

```{r}
raw_data_new  
```

```{r}

```

```{r}
library(gt)
library(gapminder)
```

### Results

Creating tables and figures or graphs showing the relationship between fecal contamination at water points and other factors such as the installer, the country, and the management system.

```{r}
#| label: tbl-country-fecal
#| tbl-cap: "Presence and absence counts in Uganda"

summary_table_country <- raw_data_new |>
  count(country_id, fecal_presence)

summary_table_country |> 
  gt() |> 
   fmt_number(columns = c(country_id, fecal_presence),
             decimals = 1)
```

@tbl-country_fecal shows the number of water sources with presence of fecal contamination.

```{r}
#| label: fig-country-fecal
#| fig-cap: "Graph showing presence and absence counts in Uganda" 
ggplot(summary_table_country, aes(x = country_id, y = n,fill = fecal_presence)) +
  geom_col(position = "stack")+
  labs(title = "Presence and absence counts in Uganda", x = "Country", y = "number" ) +
  theme_minimal()
```

@fig-country-fecal shows a graph with the presence and absence of fecal contamination at water sources.

```{r}
raw_data_new <- raw_data_new |> 
  mutate(management = case_when(
    management == "Community Management" ~ "CM",
    management == "Other" ~ "Ot",
    management == "Other Institutional Management" ~ "OIM",
    management == "Private Operator/ Delegated Management" ~ "PO/DM",
    management == "NA" ~ "NA"
  ))
```

```{r}
#| label: tbl-management-fecal
#| tbl-cap: "Presence and absence counts in management systems"

summary_table_management <- raw_data_new |>
 count(management, fecal_presence)

summary_table_management |> 
  gt() |> 
   fmt_number(columns = c(management, fecal_presence),
             decimals = 1)
```

@tbl-management-fecal shows the different management systems and the presence of fecal contamination.

```{r}
#| label: fig-management-fecal
#| fig-cap: "Graph showing presence and absence counts in management systems" 
ggplot(summary_table_management, aes(x = management, y = n,fill = fecal_presence)) +
  geom_col(position = "stack")+
  labs(title = "Presence and absence counts in different managements", x = "management system", y = "number" ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

@fig-management-fecal shows a graph that depicts results in the table.

```{r}
#| label: tbl-installer-fecal
#| tbl-cap: "Presence and absence counts in water sources by different installers"

summary_table_installer <- raw_data_new |>
  count(installer, fecal_presence)

 summary_table_installer |> 
  gt() |> 
   fmt_number(columns = c(installer, fecal_presence),
             decimals = 1)
```

@tbl-installer-fecal shows the fecal presence where different organisations have installed water points.

```{r}
#| label: fig-installer-fecal
#| fig-cap: "Graph showing presence and absence counts in water sources by different installers" 
ggplot(summary_table_installer, aes(x = installer, y = n,fill = fecal_presence)) +
  geom_col(position = "stack")+
  labs(title = "Presence and absence counts in water sources by different installers", x = "installer", y = "number" ) +
  theme_minimal()
```

@fig-installer-fecal shows fecal contamination in a graph format.

### Conclusions

Fecal contamination in water sources is existent where communities are in chage of managing water sources.

It is also prominent where other organisations that have not been recognised are in charge of constructing water points.

Uganda has been seen to have the highest water sources with fecal contamination and others are unknown.

The study suggested that improving sanitary completion and local environmental hygiene was more important than controlling on-site sanitation in improving the quality of these springs.

### References

[@howard2003; @pedley1997; @powell2003]
